Towards a free/open-source universal-dependency treebank for Kazakh

### 2015-06-20 2015-turklang-kazdep/paper.tex
better
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60878 72bbbca6-d526-0410-a7d9-f06f51895060
citation hargling
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60879 72bbbca6-d526-0410-a7d9-f06f51895060

%\usepackage{times}
\setmainfont{Times New Roman}
\newfontfamily\udfont[Scale=1.3,Letters=SmallCaps]{Linux Libertine O}

@@

\usepackage{natbib}
\bibdata{paper}

%%%%% Stuff for this paper %%%%%
\usepackage{tikz-dependency}
\newcommand{\gmk}[1]{\texttt{#1}}
\newcommand{\gloss}[1]{#1}
\newcommand{\tag}[1]{\texttt{‹#1›}}
%\newcommand{\udtag}[1]{\texttt{@#1}}
\newcommand{\udtag}[1]{\textsc{\texttt{#1}}}
\newcommand{\udlabel}[1]{{\udfont #1}}
%\newcommand{\corrauthor}[1]{\footnotemark\footnotetext{hargle}}

@@

\citet{Lynn12}, \citet{Atalay03}, \citet{Oflazer03}, \citet{DeMarneffe14}

@@

Kazakh (қазақ тілі), a Turkic language of Central Asia with around 12 million speakers, hargle bargle. %FIXME

In the process of designing a Kazakh transducer \citep{Washington14}, several decisions were made about how various inflectional morphemes affect the part of speech and syntactic function of various word forms.  These decisions include the relations to other words defined by ``case'' morphemes, and processes of ``zero derivation''.

@@

Preprocessing the corpus consists of running the text through the Kazakh morphological
analyser \citep{Washington14}, which also performs tokenisation of multiword units based 
the longest match left-to-right. Tokenisation for Kazakh is a non-trivial task, and so
we do not simply take space as a delimiter. The morphological analyser returns all 
the possible morphological analyses for each word based on a lexicon of around 20,000 lexemes.
After tokenisation and morphological analysis, the text is processed with a constraint-grammar 
based disambiguator for Kazakh consisting of 113 rules which remove inappropriate 
analyses in context. This reduces the average number of analyses per word from around 3.4
to around 1.7.

@@

In order to test the utility of the treebank in a real-world setting, we trained
and evaluated a number of models using the popular MaltParser tool \citep{nivre07}.
MaltParser is a toolkit for data-driven dependency parsing, it can learn a parsing
model from treebank data and apply this model to parse unseen sentences. The parser
has a large number of options and parameters that need to be optimised.
To select the best parser configuration we relied on
MaltOptimiser \citep{ballesteros15}. The optimiser recommended the \emph{covnonproj}
parsing algorithm.

@@

\bibliographystyle{apalike}
\bibliography{paper}

>>>>

%\usepackage{times}
\setmainfont{Times New Roman}
\newfontfamily\ll[]{Linux Libertine O}
\newfontfamily\udfont[Scale=1.3,Letters=SmallCaps]{Linux Libertine O}

@@

%\usepackage{natbib}
%\bibdata{paper}
%\usepackage[backend=bibtex]{biblatex}
\usepackage[backend=biber,citestyle=apa,style=apa]{biblatex}
\bibliography{paper}

%%%%% Stuff for this paper %%%%%
\usepackage{tikz-dependency}
\newcommand{\gmk}[1]{\texttt{#1}}
\newcommand{\gloss}[1]{#1}
\newcommand{\tag}[1]{\texttt{‹#1›}}
%\newcommand{\udtag}[1]{\texttt{@#1}}
\newcommand{\udtag}[1]{{\ll \textsc{#1}}}
\newcommand{\udlabel}[1]{{\udfont #1}}
%\newcommand{\corrauthor}[1]{\footnotemark\footnotetext{hargle}}

@@

\textcite{Lynn12}, \textcite{Atalay03}, \textcite{Oflazer03}, \textcite{DeMarneffe14}

@@

Kazakh (қазақ тілі), a Turkic language of Central Asia with around 13 million speakers \parencite{ethnologue}, hargle bargle. %FIXME

While works like \textcite{Balaqayev54} provide decent syntactic descriptions, there is little to no work on the syntax of Kazakh within modern theoretical syntactic frameworks (though work on related languages, such as Turkish, exists).

In the process of designing a Kazakh transducer \parencite{Washington14}, several decisions were made about how various inflectional morphemes affect the part of speech and syntactic function of various word forms.  These decisions include the relations to other words defined by ``case'' morphemes, and processes of ``zero derivation''.

@@

Preprocessing the corpus consists of running the text through the Kazakh morphological
analyser \parencite{Washington14}, which also performs tokenisation of multiword units based 
the longest match left-to-right. Tokenisation for Kazakh is a non-trivial task, and so
we do not simply take space as a delimiter. The morphological analyser returns all 
the possible morphological analyses for each word based on a lexicon of around 20,000 lexemes.
After tokenisation and morphological analysis, the text is processed with a constraint-grammar 
based disambiguator for Kazakh consisting of 113 rules which remove inappropriate 
analyses in context. This reduces the average number of analyses per word from around 3.4
to around 1.7.

@@

In order to test the utility of the treebank in a real-world setting, we trained
and evaluated a number of models using the popular MaltParser tool \parencite{nivre07}.
MaltParser is a toolkit for data-driven dependency parsing, it can learn a parsing
model from treebank data and apply this model to parse unseen sentences. The parser
has a large number of options and parameters that need to be optimised.
To select the best parser configuration we relied on
MaltOptimiser \parencite{ballesteros15}. The optimiser recommended the \emph{covnonproj}
parsing algorithm.

@@

%\bibliographystyle{apalike}
%\bibliography{paper}
\printbibliography
